Well-quasi-orderings on word languages

### 2025-10-01 src/conclusion.tex
fix: knowledges + refs + trying to prove uur decidability for pure morphic words

\AP It is the case that \kl{bounded languages}, studied in
\cref{infixes-bounded:sec}, have a polynomial factor complexity: such a
language is by definition included in $w_1^* \cdots w_n^*$ for some words $w_1,
\dots, w_n$, and an infix is a choice of the number of occurrences of each
$w_i$ in the infix, plus a choice of beginning and ending of the infix in the
leftmost and rightmost $w_i$. We showed in \cref{bounded-language:thm} and
particularly in \cref{bounded-language:lem}, that \kl{bounded languages} that
are \kl{well-quasi-ordered} for the \kl{infix relation} are in fact included in
some finite union of languages of the form $w_1^* w_2 w_3^*$, hence have
actually a quadratic factor complexity. However,
\cref{exponential-factor-complexity:lem} shows that there are
\kl{downwards-closed} languages that are \kl{well-quasi-ordered} for the
\kl{infix relation} but have an \kl{exponential factor complexity}.

@@

Our understanding of the situation is that whenever a computational model is
fixed to represent languages that are \kl{well-quasi-ordered} for the \kl{infix
relation}, the \kl{factor complexity} of the represented languages drops to a
quadratic one. This is the case for languages recognized by \kl{amalgamation
systems} \cref{amalgamation-systems:subsec}, that are \kl{bounded languages}
when they are \kl{well-quasi-ordered} (\cref{infix-amalgamation:thm}). 
This is
also the case for languages described as the infixes of a finite set of pairs
of \kl{morphic sequences}. Indeed, the factor complexity of a \kl{morphic
sequence} that is \kl{uniformly recurrent} is linear \cite[Theorem 24]{NIPR09}.
and therefore the factor complexity of the set of infixes of a bi-infinite word
obtained as $\bar{w_1} u w_2$ where $w_1, w_2$ are \kl{morphic sequences} and
$u$ is a finite word is quadratic.

>>>>

\AP It is the case that \kl{bounded languages}, studied in
\cref{infixes-bounded:sec}, have a polynomial factor complexity: such a
language is by definition included in $w_1^* \cdots w_n^*$ for some words $w_1,
\dots, w_n$, and an infix is a choice of the number of occurrences of each
$w_i$ in the infix, plus a choice of beginning and ending of the infix in the
leftmost and rightmost $w_i$. We showed in \cref{bounded-language:thm} and
particularly in \cref{bounded-language:lem}, that \kl{bounded languages} that
are \kl{well-quasi-ordered} for the \kl{infix relation} are in fact included in
some finite union of languages of the form $w_1^* w_2 w_3^*$, hence have
actually a quadratic factor complexity. However,
\cref{exponential-factor-complexity:lem} shows that there are
\kl{downwards closed} languages that are \kl{well-quasi-ordered} for the
\kl{infix relation} but have an \kl{exponential factor complexity}.

@@

Our understanding of the situation is that whenever a computational model is
fixed to represent languages that are \kl{well-quasi-ordered} for the \kl{infix
relation}, the \kl{factor complexity} of the represented languages drops to a
quadratic one. This is the case for languages recognized by \kl{amalgamation
systems} \cref{amalgamation-systems:subsec}, that are \kl{bounded languages}
when they are \kl{well-quasi-ordered} (\cref{infix-amalgamation:thm}). This is
also the case for languages described as the infixes of a finite set of pairs
of \kl{morphic sequences}. Indeed, the factor complexity of a \kl{morphic
sequence} that is \kl{uniformly recurrent} is linear \cite[Theorem 24]{NIPR09},
therefore the \kl{factor complexity} of a language given by \kl{sequence
representation} using \kl{morphic sequences} is at most quadratic.

### 2025-10-01 src/infixes.tex
fix: knowledges + refs + trying to prove uur decidability for pure morphic words

\AP Given an effective representation of sequences, one obtains an effective
representation of languages via \kl{sequence representations}. We can leverage
the theory of \kl{automatic sequences}, that is tailored to study the
regularity of infinite words to build a decision procedure for
\kl{well-quasi-ordered} languages for the \kl{infix relation}. A sequence $w
\in \Sigma^\Nat$ is \intro(sequence){automatic} whenever there exists a base $b
\in \Nat$ and a finite automaton that can compute the $i$-th letter of $w$
given as input the number $i$ in base $b$. The \kl{Thue-Morse sequence} is an
example of an \kl{automatic sequence}.

@@

Based on \cref{automatic-uur:lem}, we can now prove the following theorem.
\begin{theorem}
    \label{automatic-wqo:thm}
    Given a \kl{sequence representation} of a language $L \subseteq
    \Sigma^*$ where all infinite words are \kl{automatic sequences}, one can
    decide whether $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{theorem}
\begin{proof}
    It is easy to see that $L$ is \kl{well-quasi-ordered} for the \kl{infix
    relation} if and only if for every triple $(w_i^-, a_i, w_i^+)$ in the
    \kl{sequence representation} of $L$, the (potentially bi-infinite) word
    $\mathsf{reversed}(w_i^-) a_i w_i^+$ defines a \kl{well-quasi-ordered}
    language. By \cref{from-bi-to-single:lem}, this is the case if and only if
    both $w_i^-$ and $w_i^+$ are \kl{ultimately uniformly recurrent}. 
    Since one can decide whether an \kl{automatic sequence} is
    \kl{ultimately uniformly recurrent} from \cref{automatic-uur:lem},
    we conclude the proof.
\end{proof}

\todo[inline]{properly introduce morphic sequences, and maybe talk about
them in the introduction}

\AP
For \kl{automatic sequences}, the question of being \kl{ultimately uniformly
recurrent} is easily seen to be decidable, but symbolic dynamics provides a
more general framework to study infinite words. A \intro{morphic sequence} is
obtained by iterating a \kl(substitution){substitution} on a finite alphabet,
and then applying a \kl(homomorphism){homomorphism}. The \kl{Fibonacci
sequence} is an example of a \kl(morphic){morphic sequence}.
It was recently shown that one can decide whether a \kl(morphic){morphic
sequence} is \kl{uniformly recurrent} \cite[Theorem 1]{DURAND13}.
We can leverage this result to obtain the following lemma.

\begin{lemma}
  \label{morphic-uur:lem}
    Given a \kl(morphic){morphic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}

From \cref{morphic-uur:lem}, we can now prove the following corollary,
which is a generalization of \cref{automatic-wqo:thm}.

\begin{corollary}
    \label{morphic-wqo:cor}
    Given a \kl{sequence representation} of a language $L \subseteq
    \Sigma^*$ where all infinite words are \kl(morphic){morphic sequences}, one can
    decide whether $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{corollary}

>>>>

\AP Given an effective representation of sequences, one obtains an effective
representation of languages via \kl{sequence representations}. In this section,
we will rely on definitions originating from the area of symbolic dynamics,
that precisely study infinite words whose generation follows from a finitely
described process. However, we will not assume that the reader is familiar with
this domain, and we will use as black-boxes key results from this area.

\AP A first model that one can use to represent infinite words is the model of
\intro{automatic sequences}. In this case, the infinite word $w$ is described
by a finite state automaton, that can compute the $i$-th letter of the word $w$
given as input the number $i$ written in some base $b \in \Nat$. An example of
such a sequence is the \kl{Thue-Morse sequence} that can be described by a
finite automaton using a binary representation of the indices. The good
algorithmic properties of \kl{automatic sequences} come from the fact that a
Presburger definable property that uses letters of the sequence can be
(trivially) translated into a finite automaton that reads the base $b$
representation of the free variables (that are indices of the sequence). We
briefly explain in the following lemma the folklore result that one can decide
if an \kl{automatic sequence} is \kl{ultimately uniformly recurrent}.

@@

Based on \cref{automatic-uur:lem}, we can now prove the following theorem.
\begin{theorem}
    \label{automatic-wqo:thm}
    Given a \kl{sequence representation} of a language $L \subseteq
    \Sigma^*$ where all infinite words are \kl{automatic sequences}, one can
    decide whether $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{theorem}
\begin{proof}
    It is easy to see that $L$ is \kl{well-quasi-ordered} for the \kl{infix
    relation} if and only if for every triple $(w_i^-, a_i, w_i^+)$ in the
    \kl{sequence representation} of $L$, the (potentially bi-infinite) word
    $\mathsf{reversed}(w_i^-) a_i w_i^+$ defines a \kl{well-quasi-ordered}
    language. By \cref{from-bi-to-single:lem}, this is the case if and only if
    both $w_i^-$ and $w_i^+$ are \kl{ultimately uniformly recurrent}. 
    Since one can decide whether an \kl{automatic sequence} is
    \kl{ultimately uniformly recurrent} using \cref{automatic-uur:lem},
    we conclude the proof.
\end{proof}

\AP In fact, \kl{automatic sequences} are part of a larger family of sequences
studied in symbolic dynamics, called \kl{morphic sequences}. Let us first
recall that a \intro(words){morphism} is a function $f \colon \Sigma^* \to
\Gamma^*$ such that for every $u,v \in \Sigma^*$, $f(uv) = f(u)f(v)$. A
\intro{morphic sequence} $w$ is an infinite word obtained by iterating a
\kl(words){morphism} $f \colon \Sigma^* \to \Sigma^*$ on a letter $a \in
\Sigma$ such that $f(a)$ starts with $a$, and then applying a
\kl(words){homomorphism} $h \colon \Sigma^* \to \Gamma^*$. The infinite word
$f^\omega(a)$ is the limit of the sequence $\seqof{f^n(a)}[n \in \Nat]$, which
is well-defined because $f(a)$ starts with $a$, and the \kl{morphic sequence}
is $ w \defined h(f^\omega(a))$.

\AP
Every \kl{automatic
sequence} is a \kl{morphic sequence}, but not the other way around. We refer
the reader to a short survey of \cite{ALSZ17} for more details on the possible
variations on the definition of \kl{morphic sequences} and their relationships.
It was relatively recently proven that one can decide whether a \kl{morphic
sequence} is \kl{uniformly recurrent} \cite[Theorem 1]{DURAND13}. We were not
able to find in the literature whether one can decide \kl{ultimate uniform
recurrence}, but we can prove it for a restricted class of \kl{morphic
sequences}, that are called \intro{pure morphic sequences}, and are precisely
those that do not use a final \kl(words){homomorphism} $h$. Following the same proof
as for \cref{automatic-wqo:thm}, one deduces from the upcoming
\cref{morphic-uur:lem} the following \cref{morphic-wqo:cor}. We actually
conjecture that \cref{morphic-uur:lem} can be extended to all \kl{morphic
sequences} \cref{morphic-uur:conj}.

\begin{lemma}
  \label{morphic-uur:lem}
    Given a \kl{pure morphic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}
\begin{proof}
  On the one hand, one can guess a bound $N_0$ and check that 
  the suffix of the word $w_{\geq N_0}$ is \kl{uniformly recurrent},
  since the suffix of a \kl{morphic sequence} is also a \kl{morphic sequence},
  and \kl{uniform recurrence} is decidable for \kl{morphic sequences}
  \cite[Theorem 2]{DURAND13}.

  On the other hand, we can also provide an upper bound on the possible values
  of $N_0$ if $w$ is \kl{ultimately uniformly recurrent}. Let us write $f(a) =
  a s$ for some finite word $s \in \Sigma^*$. If $s$ is $f$-bounded, i.e., if
  the size of $f^n(s)$ is bounded by some constant $K \in \Nat$, then $w$ is
  ultimately periodic, and the prefix $N_0$ that may not be periodic can easily
  be bounded by looking at the morphism $f$. Let us consider the case when $s$
  is not $f$-bounded. If the word $w$ is \kl{ultimately uniformly recurrent},
  then there exists a bound $N_0$ such that $w_{\geq N_0}$ is \kl{uniformly
  recurrent}. In particular, for a large enough $n$, there exists $k_0$  such
  that for all $k \geq k_0$, the word $f^n(s)$ is an \kl{infix} of $f^{n +
  k}(s)$.
  \todo[inline]{not clear how to conclude ...}

\end{proof}

\begin{corollary}
    \label{morphic-wqo:cor}
    Given a \kl{sequence representation} of a language $L \subseteq
    \Sigma^*$ where all infinite words are \kl{pure morphic sequences}, one can
    decide whether $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{corollary}

\begin{conjecture}
  \label{morphic-uur:conj}
    Given a \kl{morphic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether it is \kl{ultimately uniformly recurrent}.
\end{conjecture}

### 2025-10-01 src/introduction.tex
fix: knowledges + refs + trying to prove uur decidability for pure morphic words

\begin{figure}[h]
    \centering
    \begin{tabular}{c|c|c|c}
        \toprule
        $L$ & \textbf{Characterisation} & $\oWidth{L}$ & $\oType{L}$ \\
        \midrule
        arbitrary & Finite unions of \kl{chains} \cref{prefixes:thm} & $< \omegaOrd$ & $< \omegaOrd^2$ \\
        regular   & Finite unions of \kl{regular} \kl{chains} \cref{prefix-wqo-reg-decidable:cor} & $< \omegaOrd$ & $< \omegaOrd^2$ \\
        \bottomrule
    \end{tabular}
    \caption{Summary of results for the \kl{prefix relation} (and symmetrically, for the \kl{suffix relation}).}
    \label{prefixes-summary:fig}
\end{figure}

@@

Turning our attention to decision procedures, we consider two computational
models respectively tailored to \kl{downwards closed} languages and to
\kl{bounded languages}. For \kl{downwards closed} languages, we consider a
model based on representations of infinite words
(\cref{decision-procedures:sec}), for which we provide a decision procedure
(\cref{automatic-wqo:thm}). The model used to represent these infinite words is
based on \kl{automatic sequences} and \kl{morphic sequences} \cite{CAKA97},
which are well-studied in the context of symbolic dynamics.
For \kl{bounded languages}, we consider the model
of \kl{amalgamation systems} \cite{ASZZ24}, which is an abstract computational
model that encompasses many classical ones, such as finite automata,
context-free grammars, and Petri nets \cite{ASZZ24}. We show that if a language
recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered} by the
\kl{infix relation}, then it is a \kl{bounded language}
(\cref{infix-amalgamation:thm}), and is therefore \kl{regular}. Furthermore, we
show that we can decide whether a given language recognized by an
\kl{amalgamation system} is \kl{well-quasi-ordered} by the \kl{infix relation}
(\cref{infix-wqo-is-emptiness:thm}). We defer the introduction of
\kl{amalgamation systems} to \cref{amalgamation-systems:subsec}.

@@

A similar question was studied by Atminas, Lozin, and Moshkov \cite{ALM17}, in
the hope of finding characterizations of classes of \emph{finite graphs} that
are \kl{well-quasi-ordered} by the \emph{induced subgraph relation}
\cite[Section 7]{ALM17}. In this setting, it is common to refer to classes of
graphs via a list of \emph{forbidden patterns}, which are finite graphs that
cannot be found as induced subgraphs in the class. Applying this reasoning to
finite words with the \kl{infix relation}, they provide an efficient decision
procedure for checking whether a language $L \subseteq \Sigma^*$ is
\kl{well-quasi-ordered} by the \kl{infix relation} whenever said language is
given as input via a list of \emph{forbidden factors} \cite[Theorem 1, Theorem
2]{ALM17}. The key construction of their paper is to study languages $L$ that
are \emph{regular} (recognized by some finite deterministic automata), for
which they can decide whether $L$ is \kl{well-quasi-ordered} by the \kl{infix
relation} \cite[Theorem 1]{ALM17}. Because it is easy to transform a list of
forbidden factors into a regular language \cite[Theorem 1]{ALM17}, this yields
the desired decision procedure. Our work extends this result in several ways:
first, we also consider the \kl{prefix relation} and the \kl{suffix relation},
then we consider non-regular languages, and finally, we provide very precise
descriptions of the \kl{well-quasi-ordered} languages, as well as tight bounds
on their \kl{ordinal invariants}. In particular, we highlight the role of
\kl{bounded languages} as the one that are well-behaved with respect to being
\kl{well-quasi-ordering} by the \kl{infix relation} \cref{infixes-bounded:sec}.
These non-trivial extensions continue to rely
on classical word combinatorial techniques that are already present in the work
of Atminas, Lozin, and Moshkov \cite[Section 3]{ALM17}.

>>>>

\begin{figure}[h]
    \centering
    \begin{tabular}{c|c|c|c}
        \toprule
        $L$ & \textbf{Characterisation} & $\oWidth{L}$ & $\oType{L}$ \\
        \midrule
        arbitrary & Finite unions of \kl{chains} \cref{prefixes:thm} & $< \omegaOrd$ & $< \omegaOrd^2$ \\
        regular   & Finite unions of \kl(language){regular} \kl{chains} \cref{prefix-wqo-reg-decidable:cor} & $< \omegaOrd$ & $< \omegaOrd^2$ \\
        \bottomrule
    \end{tabular}
    \caption{Summary of results for the \kl{prefix relation} (and symmetrically, for the \kl{suffix relation}).}
    \label{prefixes-summary:fig}
\end{figure}

@@

Turning our attention to decision procedures, we consider two computational
models respectively tailored to \kl{downwards closed} languages and to
\kl{bounded languages}. For \kl{downwards closed} languages, we consider a
model based on representations of infinite words
(\cref{decision-procedures:sec}), for which we provide a decision procedure
(\cref{automatic-wqo:thm}). The model used to represent these infinite words is
based on \kl{automatic sequences} and \kl{morphic sequences} \cite{CAKA97},
which are well-studied in the context of symbolic dynamics. For \kl{bounded
languages}, we consider the model of \kl{amalgamation systems} \cite{ASZZ24},
which is an abstract computational model that encompasses many classical ones,
such as finite automata, context-free grammars, and Petri nets \cite{ASZZ24}.
We show that if a language recognized by an \kl{amalgamation system} is
\kl{well-quasi-ordered} by the \kl{infix relation}, then it is a \kl{bounded
language} (\cref{infix-amalgamation:thm}), and is therefore
\kl(language){regular}. Furthermore, we show that we can decide whether a given
language recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered}
by the \kl{infix relation} (\cref{infix-wqo-is-emptiness:thm}). We defer the
introduction of \kl{amalgamation systems} to
\cref{amalgamation-systems:subsec}.

@@

A similar question was studied by Atminas, Lozin, and Moshkov \cite{ALM17}, in
the hope of finding characterizations of classes of \emph{finite graphs} that
are \kl{well-quasi-ordered} by the \emph{induced subgraph relation}
\cite[Section 7]{ALM17}. In this setting, it is common to refer to classes of
graphs via a list of \emph{forbidden patterns}, which are finite graphs that
cannot be found as induced subgraphs in the class. Applying this reasoning to
finite words with the \kl{infix relation}, they provide an efficient decision
procedure for checking whether a language $L \subseteq \Sigma^*$ is
\kl{well-quasi-ordered} by the \kl{infix relation} whenever said language is
given as input via a list of \emph{forbidden factors} \cite[Theorem 1, Theorem
2]{ALM17}. The key construction of their paper is to study languages $L$ that
are \emph{regular} (recognized by some finite deterministic automata), for
which they can decide whether $L$ is \kl{well-quasi-ordered} by the \kl{infix
relation} \cite[Theorem 1]{ALM17}. Because it is easy to transform a list of
forbidden factors into a regular language \cite[Theorem 1]{ALM17}, this yields
the desired decision procedure. Our work extends this result in several ways:
first, we also consider the \kl{prefix relation} and the \kl{suffix relation},
then we consider non-regular languages, and finally, we provide very precise
descriptions of the \kl{well-quasi-ordered} languages, as well as tight bounds
on their \kl{ordinal invariants}. In particular, we highlight the role of
\kl{bounded languages} as the one that are well-behaved with respect to being
\kl{well-quasi-ordering} by the \kl{infix relation} \cref{infixes-bounded:sec}.
